Generalized Bradley-Terry models

### 2025-03-21 inst/jss_hyper3_real.Rnw
typo
$p$~value not $p$-value

<<testlambdazero, cache=TRUE, include=FALSE>>=
jj <- optimize(f=function(D){f(c(lambda=1,D))}, interval=c(0.1, 0.4), maximum=TRUE)
@
%
Further, we may calculate a support function over the
$(\lambda,D)$-plane, shown in figure~\ref{contourlambdaD}.  For
example, we may test the hypothesis that $\lambda=1$ [corresponding to
  the left hand side of figure~\ref{contourlambdaD}] for which the
support is about \Sexpr{round(f(maxv)-jj$objective,2)}, far
exceeding the two units of support per degree of freedom
criteron~\citep{edwards1992}.  Alternatively, we can observe that this
lies in the tail region of its asymptotic null $\chi^2_2$ distribution
with a $p$-value of about \Sexpr{signif(pchisq(f(maxv)-jj$objective,lower.tail=FALSE,df=2),digits=2)}.

@@

The classic red bus-blue bus problem was originally stated by
\cite{mcfadden1980} although different metaphors have been employed
for the same phenomenon by \cite{tverskey1972_jss}.  The canonical problem
is as follows.  A commuter has a choice of going to work by different
transport methods.  Here, we can assume that he has possibly
strong---but unknown---views on different forms of transport which
might incorporate a random component.  If considering whether to use a
car or a bus, the colour of the bus is of no importance to him.  Such
preferences are problematic for Bradley-Terry: If our subject is
indifferent between cars and buses, then $p_{C} = p_{RB}$.  If he is
also indifferent to colour of bus, so $p_{RB}=p_{BB}$, the following
paradox occurs.  Bradley-Terry strengths must be equal so $p_{C} =
p_{RB} = p_{BB}=\frac{1}{3}$.  Thus $\Prob(C| \left\lbrace
C,RB\right\rbrace)=\frac{1}{2}$.  Now we introduce a blue bus as an
option.  Any realistic choice model will have $\Prob(C | \left\lbrace
C,RB,BB\right\rbrace)=\frac{1}{2}$ but Bradley-Terry gives us
$\frac{1}{3}$.  Various solutions to this problem have been proposed.
\cite{tverskey1972}, for example, presents an approach involving
overlapping ``aspects'' of the elements of choice set;
\cite{mcfadden1980} suggests using a generalized extreme-value model.

@@

Specifically, it seems reasonable to assume that if we ask our
respondent to rank his options, it is highly probable that he will put
RB and BB consecutively (because they are essentially
indistinguishable).  Can we quantify the strength of this effect?  To
do this, we define a bespoke function \fct{RB\_BB\_LF} which
returns a \code{hyper3} log-likelihood function corresponding to
repeated observations of our commuter's reported ranks for the five
options.
%
<<defineredbuslf>>=
`RB_BB_LF` <- function(lambda){
    ec <- c(C=1, T=2, RB=3, BB=3, W=4) 
    h <- c(1, 1, lambda, 1)           
    (
        cheering3(v=c("RB", "BB", "C" , "T", "W"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "RB", "T" , "C", "W"), e = ec, h = h) * 2 + 
        cheering3(v=c("T" , "BB", "RB", "C", "W"), e = ec, h = h) * 5 + 
        cheering3(v=c("W" , "BB", "RB", "T", "C"), e = ec, h = h) * 4 + 
        cheering3(v=c("C" , "RB", "BB", "W", "T"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "C" , "RB", "T", "W"), e = ec, h = h) * 3
    )
}
@
%
Above, we see from the function body that he reported $RB\succ BB\succ
C\succ T\succ W$ three times [first row], and so on; perhaps his
ranking depends on the weather or how tired he is on any given day.
Vector \code{ec} places the options into equivalence classes;
equivalence class 3 comprises the two buses and the other classes are
unsupported as they are of size one.  Vector \code{h} encodes the
strength of the nonindependence.  Observe that in almost every case he
ranks RB and BB consecutively.  Function \fct{RB\_BB\_LF} takes
argument \code{lambda} that quantifies the perceived similarity
between RB and BB in the same way as function \fct{oldfirm} above.
%
<<definefunctiono>>=
o <- function(lambda){maxp(RB_BB_LF(lambda), give=1)$`Log-likelihood`} 
@ 
%
Above, we define function \fct{o} which returns the support for a
given value of $\lambda$ at the evaluate [over permissible Bradley
  Terry strengths for the options].  We can treat \fct{o} as a
support function for $\lambda$, and then maximize it:
%
<<label=maxlikesim, cache=TRUE>>=
(osup <- optimize(o, c(10, 40), maximum=TRUE))
@ 
%
So a likelihood ratio test of the null that $\lambda=1$ would be:
%
<<label=LRT_RB, cache=TRUE>>=
(suppdiff <- osup$objective - o(lambda = 1))
@
%
We can thus improve the hypothesis that $\lambda=1$, corresponding to
an unaltered Plackett-Luce system, by over 5.6 units of support by
dint of choosing instead $\lambda = 10.49$.  This easily satisfies
\citeauthor{edwards1992}'s two-units-of-support criterion;
alternatively, we could observe that this observation lies in the tail
region of its asymptotic $\chi^2_1$ distribution (Wilks's theorem),
with a $p$-value of \Sexpr{signif(pchisq(suppdiff*2,df=1,lower.tail=FALSE),3)}.
%
%It is straightforward to find a profile
%support function for $\lambda$, Figure~\ref{profilesupportlambdaRBBB}.
%We see that support of $\log\lambda$ is closely quadratic over a broad
%range.
%
%\begin{figure}[htbp]
%\begin{center}
<<plotproflike, echo=FALSE>>=
if(FALSE){
    lambda <- exp(seq(from=log(0.9), to=log(67), len=19)) 
    L <- sapply(lambda, o)
    plot(log(lambda), L-max(L), type="b")
    abline(h=c(0, -2))
    abline(v=0)
}
@ 
%\caption{A\label{profilesupportlambdaRBBB} profile support function
%  for the red bus, blue bus parameter $\lambda$, synthetic dataset.
%  Vertical line indicates null of $\lambda=1$, horizontal lines show
%  maximum support and $-2$}
%\end{center}
%\end{figure}

@@

We now apply the ideas above to a dataset obtained from the {\tt
  preflib} library~\citep{MaWa13a} in which ten different types of
sushi were placed in order of preference by 5000 judges.  Three of the
sushis (maguro, toro, tekka-maki) were made with tuna, and it is
reasonable to view these three types as being exchangeable in much the
same way as the red and blue buses.  To proceed, we need to define
equivalence classes of sushi---one for tuna-based sushis and one for
everything else.  Of course, one can also imagine the opposite
phenomenon whereby the judge decides he has had enough tuna for the
moment and this disincentivises further tuna choices.  This mechanism
would be modelled by having $\lambda<1$.  Implementation in
\code{hyper3} formalism is straightforward.  The dataset used is
\code{sushi\_table}, part of the \pkg{hyper2} package:
%
<<showsushitable>>=
noquote(head(sushi_table))
@
%
Thus, from the top line, we see that the first judge placed tamago
first, followed by anago and ika.  The equivalence classes are given
by
%
<<show_sushi_eq_classes>>=
sushi_eq_classes
@
%
so class 2 is the tuna group and class 1 everything else.  We can now
define a bespoke likelihood function for these observations, using
$\lambda$ as the argument:
%
<<defcalculate_sushi_H>>=
make_sushi_H <- function(lambda){
    H <- hyper3()
    for(i in seq_len(nrow(sushi_table))){
        H <- H + cheering3(sushi_table[i,], e=sushi_eq_classes, h=c(1, lambda))
    }
    return(H)
}
@
%
<<calclikesushi,echo=FALSE, cache=TRUE>>=
f <- function(lambda){
   H <- make_sushi_H(lambda)
   maxp(H,give=1)$`Log-likelihood`
}
s <- exp(seq(from = log(0.9), to = log(10), len = howmany))
L <- sapply(s,f)
@ 
%
% chunk 'findstuffsushi' takes about 10-20 minutes to run
%
<<findstuffsushi,echo=FALSE, cache=TRUE>>=
osushi <- optimize(f, c(2,4), maximum=TRUE)
@ 
%
% chunk tinysushi' takes about 5 minutes to run
%
<<tinysushi, echo=FALSE, cache=TRUE>>=
sushimaxlambda <- osushi$maximum
sushisupport <- osushi$objective - f(1)
@
%
% chunk 'findotherstuffsushi' takes about 10 minutes to run
%
<<findothersushistuff, echo=FALSE, cache=TRUE>>=
g <- function(lambda){f(lambda) - osushi$objective + 2}
sushi_lower <- uniroot(g, interval = exp(c(0,1)), tol = 0.01)$root
sushi_upper <- uniroot(g, interval = exp(c(1,2)), tol = 0.01)$root
@ 
%
(above, we note that the return value {\tt H} is the sum of successive
log-likelihood functions of each row of the table, assumed to be
independent).  Then, as before, we may define a profile likelihood
function of $\lambda$, Figure~\ref{profilesupportsushi}.  We
reject~$H_0\colon\lambda=1$ in favour of
$H_A\colon\lambda=\Sexpr{round(sushimaxlambda,2)}$ with about
$\Sexpr{round(sushisupport,2)}$ units of support.  Alternatively, we
may observe that this lies in the tail region of its asymptotic
$\chi^2_1$ distribution with a $p$-value of about
$\Sexpr{round(pchisq(sushisupport,df=1,lower.tail=FALSE),3)}$.  We can
therefore be reasonably sure that tuna sushi are somehow analagous to
red and blue buses, in the sense that all three tuna items tend to
occupy closely separated places in judges' preference ranks.

@@

\begin{figure}[htbp]
  \begin{center}
<<csgo_plotlikelam, echo=FALSE>>=
plot(log(lam), L - max(L), type='b')
segments(x0 = log(csgo_optlam),y0 = -0.5, y1 =  0.1, col='red')
segments(x0 = log(csgo_lower), y0 = -2.2, y1 = -1.8, col = 'blue')
segments(x0 = log(csgo_upper), y0 = -2.2, y1 = -1.8, col = 'blue')
abline(v = 0, col = 'gray')
abline(h = c(0, -2), col = 'gray')
@ 
\caption{A\label{profsupcounterstrike} profile support function for
  the red bus-blue bus parameter $\log\lambda$ applied to mutually
  supporting equivalence class
  $\left\lbrace\mbox{\skad},\mbox{\stew}\right\rbrace$.  Vertical gray
  line shows $H_0\colon\lambda =1$, for which the support is about
  $\Sexpr{round(csgo_support,2)}$ relative to the evaluate
  $\hat{\lambda}\simeq\Sexpr{round(csgo_optlam,3)}$ (small red bar),
  corresponding to a $p$-value of about
  $\Sexpr{round(csgo_pvalue,3)}$.  Horizontal lines show the
  two-units-of-support credible interval of
  $(\Sexpr{round(csgo_lower,3)},\Sexpr{round(csgo_upper,3)})$, marked
  in blue}
\end{center}
\end{figure}

To assess this, we follow exactly the same protocol as for the sushi
preference dataset but using the deathorder statistic from the dataset
above.  We specify two equivalence classes: one comprising \skad\ and
\stew, and one everyone else.  Figure~\ref{profsupcounterstrike} gives
a support function for $\lambda$ showing, among other things, that we
may reject the hypothesis that $\lambda=1$ in favour of $\lambda<1$ by
a likelihood ratio of about $\Sexpr{round(csgo_support,2)}$.  We
observe that this exceeds the two units of support criterion of
\cite{edwards1992} [asymptotic $p$-value
  $\simeq\Sexpr{round(csgo_pvalue,3)}$] so we may infer that there is
indeed some systematic mechanism---perhaps a positive feedback
loop---which tends to magnify differences in death statistic.

>>>>

<<testlambdazero, cache=TRUE, include=FALSE>>=
jj <- optimize(f=function(D){f(c(lambda=1,D))}, interval=c(0.1, 0.4), maximum=TRUE)
@
%
Further, we may calculate a support function over the
$(\lambda,D)$-plane, shown in figure~\ref{contourlambdaD}.  For
example, we may test the hypothesis that $\lambda=1$ [corresponding to
  the left hand side of figure~\ref{contourlambdaD}] for which the
support is about \Sexpr{round(f(maxv)-jj$objective,2)}, far
exceeding the two units of support per degree of freedom
criteron~\citep{edwards1992}.  Alternatively, we can observe that this
lies in the tail region of its asymptotic null $\chi^2_2$ distribution
with a $p$~value of about \Sexpr{signif(pchisq(f(maxv)-jj$objective,lower.tail=FALSE,df=2),digits=2)}.

@@

The classic red bus-blue bus problem was originally stated by
\cite{mcfadden1980} although different metaphors have been employed
for the same phenomenon by \cite{tverskey1972_jss}.  The canonical problem
is as follows.  A commuter has a choice of going to work by different
transport methods.  Here, we can assume that he has possibly
strong---but unknown---views on different forms of transport which
might incorporate a random component.  If considering whether to use a
car or a bus, the colour of the bus is of no importance to him.  Such
preferences are problematic for Bradley-Terry: If our subject is
indifferent between cars and buses, then $p_{C} = p_{RB}$.  If he is
also indifferent to colour of bus, so $p_{RB}=p_{BB}$, the following
paradox occurs.  Bradley-Terry strengths must be equal so $p_{C} =
p_{RB} = p_{BB}=\frac{1}{3}$.  Thus $\Prob(C| \left\lbrace
C,RB\right\rbrace)=\frac{1}{2}$.  Now we introduce a blue bus as an
option.  Any realistic choice model will have $\Prob(C | \left\lbrace
C,RB,BB\right\rbrace)=\frac{1}{2}$ but Bradley-Terry gives us
$\frac{1}{3}$.  Various solutions to this problem have been proposed.
\cite{tverskey1972_jss}, for example, presents an approach involving
overlapping ``aspects'' of the elements of choice set;
\cite{mcfadden1980} suggests using a generalized extreme-value model.

@@

Specifically, it seems reasonable to assume that if we ask our
respondent to rank his options, it is highly probable that he will put
RB and BB consecutively (because they are essentially
indistinguishable).  Can we quantify the strength of this effect?  To
do this, we define a bespoke function \fct{RB\_BB\_LF} which
returns a \code{hyper3} log-likelihood function corresponding to
repeated observations of our commuter's reported ranks for the five
options.
%
<<defineredbuslf>>=
`RB_BB_LF` <- function(lambda){
    ec <- c(C=1, T=2, RB=3, BB=3, W=4) 
    h <- c(1, 1, lambda, 1)           
    (
        cheering3(v=c("RB", "BB", "C" , "T", "W"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "RB", "T" , "C", "W"), e = ec, h = h) * 2 + 
        cheering3(v=c("T" , "BB", "RB", "C", "W"), e = ec, h = h) * 5 + 
        cheering3(v=c("W" , "BB", "RB", "T", "C"), e = ec, h = h) * 4 + 
        cheering3(v=c("C" , "RB", "BB", "W", "T"), e = ec, h = h) * 3 + 
        cheering3(v=c("BB", "C" , "RB", "T", "W"), e = ec, h = h) * 3
    )
}
@
%
Above, we see from the function body that he reported $RB\succ BB\succ
C\succ T\succ W$ three times [first row], and so on; perhaps his
ranking depends on the weather or how tired he is on any given day.
Vector \code{ec} places the options into equivalence classes;
equivalence class 3 comprises the two buses and the other classes are
unsupported as they are of size one.  Vector \code{h} encodes the
strength of the nonindependence.  Observe that in almost every case he
ranks RB and BB consecutively.  Function \fct{RB\_BB\_LF} takes
argument \code{lambda} that quantifies the perceived similarity
between RB and BB in the same way as function \fct{oldfirm} above.
%
<<definefunctiono>>=
o <- function(lambda){maxp(RB_BB_LF(lambda), give=1)$`Log-likelihood`} 
@ 
%
Above, we define function \fct{o} which returns the support for a
given value of $\lambda$ at the evaluate [over permissible Bradley
  Terry strengths for the options].  We can treat \fct{o} as a
support function for $\lambda$, and then maximize it:
%
<<label=maxlikesim, cache=TRUE>>=
(osup <- optimize(o, c(10, 40), maximum=TRUE))
@ 
%
So a likelihood ratio test of the null that $\lambda=1$ would be:
%
<<label=LRT_RB, cache=TRUE>>=
(suppdiff <- osup$objective - o(lambda = 1))
@
%
We can thus improve the hypothesis that $\lambda=1$, corresponding to
an unaltered Plackett-Luce system, by over 5.6 units of support by
dint of choosing instead $\lambda = 10.49$.  This easily satisfies
\citeauthor{edwards1992}'s two-units-of-support criterion;
alternatively, we could observe that this observation lies in the tail
region of its asymptotic $\chi^2_1$ distribution (Wilks's theorem),
with a $p$~value of \Sexpr{signif(pchisq(suppdiff*2,df=1,lower.tail=FALSE),3)}.
%
%It is straightforward to find a profile
%support function for $\lambda$, Figure~\ref{profilesupportlambdaRBBB}.
%We see that support of $\log\lambda$ is closely quadratic over a broad
%range.
%
%\begin{figure}[htbp]
%\begin{center}
<<plotproflike, echo=FALSE>>=
if(FALSE){
    lambda <- exp(seq(from=log(0.9), to=log(67), len=19)) 
    L <- sapply(lambda, o)
    plot(log(lambda), L-max(L), type="b")
    abline(h=c(0, -2))
    abline(v=0)
}
@ 
%\caption{A\label{profilesupportlambdaRBBB} profile support function
%  for the red bus, blue bus parameter $\lambda$, synthetic dataset.
%  Vertical line indicates null of $\lambda=1$, horizontal lines show
%  maximum support and $-2$}
%\end{center}
%\end{figure}

@@

We now apply the ideas above to a dataset obtained from the {\tt
  preflib} library~\citep{MaWa13a} in which ten different types of
sushi were placed in order of preference by 5000 judges.  Three of the
sushis (maguro, toro, tekka-maki) were made with tuna, and it is
reasonable to view these three types as being exchangeable in much the
same way as the red and blue buses.  To proceed, we need to define
equivalence classes of sushi---one for tuna-based sushis and one for
everything else.  Of course, one can also imagine the opposite
phenomenon whereby the judge decides he has had enough tuna for the
moment and this disincentivises further tuna choices.  This mechanism
would be modelled by having $\lambda<1$.  Implementation in
\code{hyper3} formalism is straightforward.  The dataset used is
\code{sushi\_table}, part of the \pkg{hyper2} package:
%
<<showsushitable>>=
noquote(head(sushi_table))
@
%
Thus, from the top line, we see that the first judge placed tamago
first, followed by anago and ika.  The equivalence classes are given
by
%
<<show_sushi_eq_classes>>=
sushi_eq_classes
@
%
so class 2 is the tuna group and class 1 everything else.  We can now
define a bespoke likelihood function for these observations, using
$\lambda$ as the argument:
%
<<defcalculate_sushi_H>>=
make_sushi_H <- function(lambda){
    H <- hyper3()
    for(i in seq_len(nrow(sushi_table))){
        H <- H + cheering3(sushi_table[i,], e=sushi_eq_classes, h=c(1, lambda))
    }
    return(H)
}
@
%
<<calclikesushi,echo=FALSE, cache=TRUE>>=
f <- function(lambda){
   H <- make_sushi_H(lambda)
   maxp(H,give=1)$`Log-likelihood`
}
s <- exp(seq(from = log(0.9), to = log(10), len = howmany))
L <- sapply(s,f)
@ 
%
% chunk 'findstuffsushi' takes about 10-20 minutes to run
%
<<findstuffsushi,echo=FALSE, cache=TRUE>>=
osushi <- optimize(f, c(2,4), maximum=TRUE)
@ 
%
% chunk tinysushi' takes about 5 minutes to run
%
<<tinysushi, echo=FALSE, cache=TRUE>>=
sushimaxlambda <- osushi$maximum
sushisupport <- osushi$objective - f(1)
@
%
% chunk 'findotherstuffsushi' takes about 10 minutes to run
%
<<findothersushistuff, echo=FALSE, cache=TRUE>>=
g <- function(lambda){f(lambda) - osushi$objective + 2}
sushi_lower <- uniroot(g, interval = exp(c(0,1)), tol = 0.01)$root
sushi_upper <- uniroot(g, interval = exp(c(1,2)), tol = 0.01)$root
@ 
%
(above, we note that the return value {\tt H} is the sum of successive
log-likelihood functions of each row of the table, assumed to be
independent).  Then, as before, we may define a profile likelihood
function of $\lambda$, Figure~\ref{profilesupportsushi}.  We
reject~$H_0\colon\lambda=1$ in favour of
$H_A\colon\lambda=\Sexpr{round(sushimaxlambda,2)}$ with about
$\Sexpr{round(sushisupport,2)}$ units of support.  Alternatively, we
may observe that this lies in the tail region of its asymptotic
$\chi^2_1$ distribution with a $p$~value of about
$\Sexpr{round(pchisq(sushisupport,df=1,lower.tail=FALSE),3)}$.  We can
therefore be reasonably sure that tuna sushi are somehow analagous to
red and blue buses, in the sense that all three tuna items tend to
occupy closely separated places in judges' preference ranks.

@@

\begin{figure}[htbp]
  \begin{center}
<<csgo_plotlikelam, echo=FALSE>>=
plot(log(lam), L - max(L), type='b')
segments(x0 = log(csgo_optlam),y0 = -0.5, y1 =  0.1, col='red')
segments(x0 = log(csgo_lower), y0 = -2.2, y1 = -1.8, col = 'blue')
segments(x0 = log(csgo_upper), y0 = -2.2, y1 = -1.8, col = 'blue')
abline(v = 0, col = 'gray')
abline(h = c(0, -2), col = 'gray')
@ 
\caption{A\label{profsupcounterstrike} profile support function for
  the red bus-blue bus parameter $\log\lambda$ applied to mutually
  supporting equivalence class
  $\left\lbrace\mbox{\skad},\mbox{\stew}\right\rbrace$.  Vertical gray
  line shows $H_0\colon\lambda =1$, for which the support is about
  $\Sexpr{round(csgo_support,2)}$ relative to the evaluate
  $\hat{\lambda}\simeq\Sexpr{round(csgo_optlam,3)}$ (small red bar),
  corresponding to a $p$~value of about
  $\Sexpr{round(csgo_pvalue,3)}$.  Horizontal lines show the
  two-units-of-support credible interval of
  $(\Sexpr{round(csgo_lower,3)},\Sexpr{round(csgo_upper,3)})$, marked
  in blue}
\end{center}
\end{figure}

To assess this, we follow exactly the same protocol as for the sushi
preference dataset but using the deathorder statistic from the dataset
above.  We specify two equivalence classes: one comprising \skad\ and
\stew, and one everyone else.  Figure~\ref{profsupcounterstrike} gives
a support function for $\lambda$ showing, among other things, that we
may reject the hypothesis that $\lambda=1$ in favour of $\lambda<1$ by
a likelihood ratio of about $\Sexpr{round(csgo_support,2)}$.  We
observe that this exceeds the two units of support criterion of
\cite{edwards1992} [asymptotic $p$~value
  $\simeq\Sexpr{round(csgo_pvalue,3)}$] so we may infer that there is
indeed some systematic mechanism---perhaps a positive feedback
loop---which tends to magnify differences in death statistic.
